Software architectures to integrate workflow engines in science gateways

### 2016-10-13 paper.tex
Various fixes in paper

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnotes;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

@@

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \address[label1]{}
%% \address[label2]{}

@@

\begin{abstract}
  Science gateways often rely on workflow engines to execute
  applications on distributed infrastructures. We investigate six software
  architectures commonly used to integrate workflow engines into
  science gateways. In \emph{tight integration}, the workflow engine
  shares software components with the science gateway. In
  \emph{service invocation}, the engine is isolated and invoked
  through a specific software interface. In \emph{task encapsulation},
  the engine is wrapped as a computing task executed on the
  infrastructure. In the \emph{pool model}, the engine is bundled in
  an agent that connects to a central pool to fetch and execute
  workflows. In \emph{nested workflows}, the engine is integrated as a
  child process of another engine. In \emph{workflow conversion}, the
  engine is integrated through workflow language conversion. We
  describe and evaluate these architectures with metrics for
  assessment of integration complexity, robustness, extensibility,
  scalability and \correction{functionality}. Tight integration and task encapsulation are the easiest
  to integrate and the most robust. Extensibility is equivalent in
  most architectures. The pool model is the most scalable one and
  meta-workflows are only available in task encapsulation, nested
  workflows and workflow conversion. These results provide insights
  for science gateway architects and developers.
\end{abstract}

@@

Workflow engines are critical components in the ecosystem of tools
and services offered by science gateways. Several software
architectures can be adopted to integrate workflow engines in science
gateways, with important consequences for the development effort
required and resulting system.

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experience in
the development and sustained operation of the CBRAIN~\cite{SHER-14},
VIP~\cite{GLAT-13} and NSG~\cite{shahand2015data,shahand:2012jgc}
science gateways for medical image analysis during the past 7 years,
as well as by lessons learned from several science gateway and
workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}.

\correction{The analysis presented in the paper is intended for
experts in science gateway and workflow engine design. It is an
abstraction effort to identify and evaluate the fundamental
architectural patterns that are encountered while integrating workflow
engines and science gateways. In real systems, such patterns are often
intertwined due to considerations related to the historical and
technical context of software projects. The following work helps
identify the key patterns in systems and highlight potential
improvements.}

\correction{The reminder of this section provides} background
information and definitions of workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe six
architectures that cover the main patterns that we encountered in the
various systems that we studied during the last few years. The
architectures are described within a consistent framework that
underlines the functional interactions between their main software
components. \correction{We illustrate the application of our
  architectural patterns to real systems in Section~\ref{sec:real
    systems}. In Section~\ref{sec:metrics}, we introduce metrics that
  measure integration complexity, robustness, extensibility,
  scalability and \correction{functionality}. These metrics are
  specifically designed to measure the ability of the architectures to
  address workflow-related issues commonly encountered in science
  gateways. In Section~\ref{sec:evaluation}, we evaluate the different
  architectures based on the proposed metrics.} \correction{Based on
  this evaluation, we compare and discuss the architectures in
  Section~\ref{sec:discussion} where we also highlight}
recommendations for science gateway architects and
developers. \correction{Finally, in Section~\ref{sec:application}, we
  illustrate how our evaluation framework can be used to help design
  new systems. Related work is presented in Section~\ref{sec:related}.}

@@

  model, user is exposed to the workflow topology, i.e. to the outputs

@@

\item Pool-agent: specific to the pool architecture described in
  Section~\ref{sec:pool}. This is an interaction used when agents
  retrieve work from a central pool. It covers agent registration and
  de-registration to the pool, protocols to send work from the pool to
  the agent, mechanisms to update work status, and so on. A similar
  type of interaction is used in pilot-job
  systems~\cite{turilli2015comprehensive} and other agent-based
  computing models.

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a model in a
Ruby on Rails application\footnote{\correction{Note that the
    particular case of portlets can be a bit ambiguous: when portlets
    are deployed remotely, using protocols such as Web Services for
    Remote Portlets (WSRP), they become independent software
    components according to our definition because they could be
    operated by different teams on different hardware systems. In this
    case, the system should be classified in the service invocation
    architecture since the portlets actually become Web Services. The
    same comment holds for remote object invocation.}}. The workflow
engine and the science gateway usually share a database where
applications, users and other resources are stored.  In this model,
task and data control are both initiated from the science
gateway. Interactions \texttt{b} and \texttt{c$_2$} are initiated from
the workflow engine, while \texttt{c$_1$} comes from other parts of
the science gateway, for instance a data management interface. As in
any other model, the installation of new workflows in the science
gateway (\texttt{a$_2$}) and infrastructure (\texttt{a$_1$}) is done
by an administrator, for security reasons. This is the model adopted
in the Catania Science Gateway Framework~\cite{ardizzone2012decide}
(see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE)~\cite{maddineni2012distributed}, in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

This architecture is largely adopted in systems such as Apache
Airvata~\cite{marru2011apache}, GridSFEA~\cite{Elts2010}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the systems
in~\cite{wu2010accelerating,shahand:2012jgc} and the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012}. The integration between the HubZero
science gateway and the Pegasus workflow engine performed
in~\cite{CPE:CPE3257} also uses a service architecture, where the
\texttt{d} interaction is implemented as a set of calls to Pegasus'
command-line tools (e.g., pegasus-status, pegasus-plan,
etc.).

@@

Nested workflows are abstract architectural patterns that can be
instantiated in the various architectures described previously. We
focus on instantiation with the service invocation model
(Figure~\subref*{archi:nested}-Right) as this is the most used
architecture. \correction{We call this instantiation
  ``Service+Service'' since both engines are integrated through
  service invocation.} We assume that the parent
and child workflow engines are distinct pieces of software that
require different workflow services invoked by distinct \texttt{d}
interactions. If this is not the case, then workflow services can be
collapsed into a single one with a \texttt{d} interaction with
itself. An example of such interaction is the use of hierarchical
workflows in Pegasus~\cite{Deelman201517}.  Workflow engines
communicate with infrastructures using \texttt{b} and
\texttt{c$_2$}. Science gateway transfers user data to infrastructures
using \texttt{c$_1$} interactions.

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow tasks as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform.

@@

\correction{\section{Architecture of real systems}}

@@

%\todo{Introduce this section and take the text out of the captions}

\correction{This section illustrates the architectural patterns on a
  few real systems. The architecture diagrams presented are the
  original ones as published by the system developers, which we
  annotated using the interactions and components in
  Figure~\ref{fig:architectures}, in red color. Application porting
  interactions \texttt{a$_1$} and \texttt{a$_2$} are not represented
  on these diagrams.

@@

The architecture of the Catania Science Gateway Framework (CSGF) is
shown in Figure~\ref{fig:csgf}. In CSGF, workflows are integrated as
Liferay portlets, i.e., as part of the Science Gateway when no
specific protocol is used for remote invocation. A specific portlet
called \texttt{mi-parallel-portlet} allows to create job collections,
parametric jobs and split-and-merge types of workflows. The science
gateway also includes a set of services to manage data and tasks
(jobs) on a wide array of grid middleware systems through the SAGA
API.
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/CSGF}
\caption{\correction{Architecture of the Catania Science Gateway Framework (tight
  integration). Figure adapted from~\cite{ardizzone2012decide} with
  permission of the first author} \todo{approval still pending}.}
\label{fig:csgf}
\end{figure}

@@

Figure~\ref{fig:vip-architecture} shows the architecture of the
Virtual Imaging Platform, where the MOTEUR workflow
engine~\cite{GLAT-08i} is invoked through a service. The interactions
in the VIP architecture map to the ones defined by our framework in a
straightforward manner, except for interaction \texttt{c$_2$}
(management of workflow data) which, in VIP, is triggered by the
workflow tasks instead of the workflow engine. However, this does not
contradict the service invocation architecture since the tasks are
generated by the workflow engine.
\begin{figure*}
%\def\svgwidth{1.5\columnwidth}
\includegraphics[width=0.9\textwidth]{figures/VIP.eps}
\caption{Architecture of the Virtual Imaging Platform (service
  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds
  to interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP,
  workflow data transfers are performed by the tasks and embedded in
  their descriptions).}
\label{fig:vip-architecture}
\end{figure*}

@@

Task encapsulation is used in the CBRAIN gateway~\cite{SHER-14} to
integrate the FSL toolkit~\cite{Jenkinson2012782} and the PSOM
workflow engine~\cite{bellec2012pipeline}. The CBRAIN-FSL integration,
shown in Figure~\subref*{fig:cbrain-fsl-architecture}, allows
leveraging FSL pipelines written in low-level workflow languages
(Linux executables and scripts) that submit tasks uniformly through a
specific tool called fsl\_sub. The science gateway is actually split
in two services, the CBRAIN portal and the CBRAIN execution server,
deployed on a head node of the infrastructure. The mapping of
interactions \texttt{b}, \texttt{c$_1$} and \texttt{e} to the
CBRAIN-FSL interactions is straightforward.

@@

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\includegraphics[width=\columnwidth]{figures/CBRAIN-FSL.eps}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
reproduced from~\cite{GLAT-16}.]{
  \includegraphics[width=\columnwidth]{figures/CBRAIN-PSOM.eps}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (task encapsulation).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler (FSL sub-tasks or PSOM workers). FSL sub-tasks will run locally
  instead of being submitted again to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures}(III) is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

@@

Figure~\ref{fig:tavaxy} shows the architecture of Tavaxy where
workflows in various formats (tSCUFL, SCUFL, t2flow and JSON) can be
imported. As detailed in~\cite{Abouelhoda2012}, workflow conversion
(interaction \texttt{g}) is implemented through an interactive
authoring module where users can check and edit workflow imports, and
a mapper module that performs some language substitutions. The
workflow engine is a re-engineered version of the Galaxy engine that
also includes some components of Taverna.
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/Tavaxy.eps}
\caption{\correction{Architecture of Tavaxy (workflow conversion). Figure adapted
  from~\cite{Abouelhoda2012} with permission of the first author} \todo{approval is still pending}.}
\label{fig:tavaxy}
\end{figure}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation} using five main criteria:
integration complexity, robustness, extensibility, scalability and
\correction{functionality}. These are based on science gateway properties
introduced in Section~\ref{sec:intro_sg}. Criteria break down into
specific metrics where \emph{lower value indicates better
  performance}. Colors in Table~\ref{table:evaluation} represent
normalized scores (see caption). For each criterion, a global score is
computed by summing up the individual scores for each metric. We
ensure that the different metrics used to calculate the global score
for a criterion are comparable, so that they can be combined
meaningfully.

\correction{Note that the metrics defined hereafter aim at providing a framework
to evaluate architectures independently from any particular system,
regardless of any technical constraint such as the use of a specific
programming language or the internal architecture of the science
gateway. The goal is to abstract architectural discussions from the
particular technical context of a real system. Some degree of
system-specificity can be introduced by assigning weights to
interactions and components through a public spreadsheet available at
\url{https://frama.link/LrPRsCQp}. In particular, sheet "Components
and interactions" contains the weight given to these entities in the
metrics and can be used to produce new versions of
Table~\ref{table:evaluation} that fit the particular technical context
of a real system. For instance, a high weight could be assigned to the
science gateway if its internal architecture is such that it can
hardly be extended in practice.}

\correction{In addition, all the metrics in a criterion contribute
  equally to the global score used for the criterion. Column B of
  sheet ``Metrics'' of the public spreadsheet may be used to assign
  different weights to each metric to adjust the specific context of a
  study.}

@@

\correction{Integration complexity} is measured through the number of
software components and their interactions. It breaks down into the
following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total components (\texttt{I$_1$}): total number of components in
  the architecture.
\item Total interactions (\texttt{I$_2$}): total number of
  interactions in the architecture.
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not
developed nor maintained by the groups who integrate workflow engines
in science gateway. We support its inclusion here because integrating
an infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.),
therefore increasing complexity. \correction{We assume that software
  components contribute equally to $I_1$ and that interactions also
  contribute equally to $I_2$. In a real system, this may not be the
  case due to a range of project-specific practical reasons that
  cannot be covered in this paper. The weights given to components and
  interactions can be dynamically changed in the public spreadsheet
  (sheet ``Components and interactions'') to generate new versions of
  Table~\ref{table:evaluation} as required.} The two metrics
\correction{I$_1$ and I$_2$} are summed to obtain a global score that
measures the total number of software pieces affected by integration.

@@

\correction{Functionality includes}:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{\correction{F}$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{\correction{F}$_1$}=0 means that meta-workflows are
  intrinsically enabled by the architecture (i.e. they can be
  implemented using the components and interactions already in place),
  and \texttt{\correction{F}$_1$}=1 means that they may be implemented with some
  development effort.

@@

\section{\correction{Architecture evaluation}}

@@

%\todo{Integrating a workflow engine as a sub-task
%only requires to implement interaction \texttt{e} (\texttt{I$_2$}=1)
%and to install the workflow engine on the infrastructure
%(\texttt{I$_3$}=1). A new task type is created in the science gateway
%only when a new workflow has to be integrated (see M$_4$), but this is
%not required to integrate the engine itself (\texttt{I$_1$}=0).}

@@

Differences in \correction{functionality} should not be neglected. Task
encapsulation, nested workflows (with service invocation) and workflow
conversion (with service invocation) are the only architectures that
intrinsically support meta-workflows. Besides, tight integration is
the only architecture that allows fine-grained debugging, which may be
critical for efficient user support.

Table~\ref{table:overall} \todo{update with new architecture?} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
Note that this analysis is only meant for illustration purposes, since
it assumes that all criteria have equal weights while real systems
would favor some of them. \correction{Weights can be adjusted in the
  public spreadsheet at \url{https://frama.link/LrPRsCQp} (column B of
  sheet ``Metrics'') to produce a custom version of
  Table~\ref{table:overall} if required.}

>>>>

@@

@@

\begin{abstract}
  Science gateways often rely on workflow engines to execute
  applications on distributed infrastructures. We investigate six
  software architectures commonly used to integrate workflow engines
  into science gateways. In \emph{tight integration}, the workflow
  engine shares software components with the science gateway. In
  \emph{service invocation}, the engine is isolated and invoked
  through a specific software interface. In \emph{task encapsulation},
  the engine is wrapped as a computing task executed on the
  infrastructure. In the \emph{pool model}, the engine is bundled in
  an agent that connects to a central pool to fetch and execute
  workflows. In \emph{nested workflows}, the engine is integrated as a
  child process of another engine. In \emph{workflow conversion}, the
  engine is integrated through workflow language conversion. We
  describe and evaluate these architectures with metrics for
  assessment of integration complexity, robustness, extensibility,
  scalability and \correction{functionality}. Tight integration and
  task encapsulation are the easiest to integrate and the most
  robust. Extensibility is equivalent in most architectures. The pool
  model is the most scalable one and meta-workflows are only available
  in nested workflows and workflow conversion. These results provide
  insights for science gateway architects and developers.
\end{abstract}

@@

Several software architectures can be adopted to integrate workflow
engines in \correction{the ecosystem of tools and services offered by}
science gateways, with important consequences for the development
effort required and resulting system.

This paper describes, \correction{illustrates} and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experience in
the development and sustained operation of the CBRAIN~\cite{SHER-14},
NSG~\cite{shahand2015data,shahand:2012jgc} and VIP~\cite{GLAT-13}
science gateways during the past 7 years, as well as by lessons
learned from several science gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}.

\correction{This analysis is intended for experts in science gateway
  and workflow engine design. It is an abstraction effort to identify
  and evaluate the fundamental architectural patterns that are
  encountered while integrating workflow engines and science
  gateways. In real systems, such patterns are often intertwined and
  obfuscated due to the historical and
  technical context of software projects.}

\correction{The remainder of this section provides} background
information and definitions of workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe six
architectures within a consistent framework that underlines the
functional interactions between their main software
components. \correction{We illustrate these architectures on real
  systems in Section~\ref{sec:real systems}.} In
  Section~\ref{sec:metrics}, we introduce metrics that measure
  integration complexity, robustness, extensibility, scalability and
  \correction{functionality}. The metrics are specifically designed to
  measure the ability of the architectures to address workflow-related
  issues commonly encountered in science gateways. \correction{We
    evaluate the architectures individually in
    Section~\ref{sec:evaluation}, and we compare and discuss them in
    Section~\ref{sec:discussion}.} \correction{Finally, in
    Section~\ref{sec:application}, we illustrate how our evaluation
    framework can be used to help design new systems. Related work is
    presented in Section~\ref{sec:related}.}

@@

  model, user is exposed to the workflow topology, i.e., to the outputs

@@

\item Pool-agent: specific to the pool architecture. This is an
  interaction used when agents retrieve work from a central pool. It
  covers agent registration and de-registration to the pool, protocols
  to send work from the pool to the agent, mechanisms to update work
  status, and so on. A similar type of interaction is used in
  pilot-job systems~\cite{turilli2015comprehensive} and other
  agent-based computing models.

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a model in a
Ruby on Rails application\footnote{\correction{Note that the
    particular case of portlets can be a bit ambiguous: when portlets
    are deployed remotely, using protocols such as Web Services for
    Remote Portlets (WSRP), they become independent software
    components according to our definition because they could be
    operated by different teams on different hardware systems. In this
    case, the system should be classified in the service invocation
    architecture since the portlets actually become Web Services. The
    same comment holds for remote object invocation.}}. The workflow
engine and the science gateway usually share a database where
applications, users and other resources are stored.  In this model,
task and data control are both initiated from the science
gateway. Interactions \texttt{b} and \texttt{c$_2$} are initiated from
the workflow engine, while \texttt{c$_1$} comes from other parts of
the science gateway, for instance a data management interface. As in
any other model, the installation of new workflows in the science
gateway (\texttt{a$_2$}) and infrastructure (\texttt{a$_1$}) is done
by an administrator, for security reasons. \correction{Tight
  integration} is the model adopted in the Catania Science Gateway
Framework~\cite{ardizzone2012decide} (see specific documentation on
workflows\footnote{\url{https://frama.link/jfzgaEbj}}), in the
Distributed application runtime environment
(DARE)~\cite{maddineni2012distributed}, in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

This architecture is largely adopted in systems such as
\correction{the} Apache Airvata
\correction{framework}~\cite{marru2011apache},
GridSFEA~\cite{Elts2010}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the systems
in~\cite{wu2010accelerating,shahand:2012jgc} and the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012}. The integration between the HubZero
science gateway and the Pegasus workflow engine performed
in~\cite{CPE:CPE3257} also uses a service architecture, where the
\texttt{d} interaction is implemented as a set of calls to Pegasus'
command-line tools (e.g., pegasus-status, pegasus-plan, etc.).

@@

Nested workflows are abstract architectural patterns that can be
instantiated in the various architectures described previously. We
focus on instantiation with the service invocation model
(Figure~\subref*{archi:nested}-Right) as this is the most used
architecture. \correction{We call this instantiation
  ``Service+Service'' since both engines are integrated through
  service invocation.} We assume that the parent
and child workflow engines are distinct pieces of software that
require different workflow services invoked by distinct \texttt{d}
interactions. If this is not the case, then workflow services can be
collapsed into a single one with a \texttt{d} interaction with
itself. An example of such interaction is the use of hierarchical
workflows in Pegasus~\cite{Deelman201517}.  Workflow engines
communicate with infrastructures using \texttt{b} and
\texttt{c$_2$}. Science gateway transfers user data to infrastructures
using \texttt{c$_1$}.

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow tasks as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e., to integrate
various workflow engines in a consistent
platform.

@@

\correction{\section{Real systems}}

@@

\correction{This section illustrates the architectural patterns on a
  few real systems, using the original architecture diagrams published
  by the system developers. To improve readability, we annotated these
  diagrams with the interactions and components used in
  Figure~\ref{fig:architectures}.

@@

The architecture of the Catania Science Gateway Framework (CSGF) is
shown in Figure~\ref{fig:csgf}. In CSGF, workflows are integrated as
Liferay portlets, i.e., as part of the science gateway when no
specific protocol is used for remote invocation. A specific portlet
called \texttt{mi-parallel-portlet} allows to create job collections,
parametric jobs and split-and-merge types of workflows. The science
gateway also includes a set of services to manage data and tasks
(jobs) on a wide array of grid middleware systems through the SAGA
API.
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/CSGF}
\caption{\correction{Architecture of the Catania Science Gateway Framework (tight
  integration). Figure adapted from~\cite{ardizzone2012decide} with
  permission of the first author} \todo{approval still pending}.}
\label{fig:csgf}
\end{figure}

@@

Figure~\ref{fig:vip-architecture} shows the architecture of the
Virtual Imaging Platform (VIP), where the MOTEUR workflow
engine~\cite{GLAT-08i} is invoked through a service. The interactions
in the VIP architecture map to the ones defined by our framework in a
straightforward manner, except for interaction \texttt{c$_2$}
(management of workflow data) which, in VIP, is triggered by the
workflow tasks instead of the workflow engine. This does not
contradict the service invocation architecture since the tasks are
generated by the workflow engine.
\begin{figure*}
%\def\svgwidth{1.5\columnwidth}
\includegraphics[width=0.9\textwidth]{figures/VIP.eps}
\caption{Architecture of the Virtual Imaging Platform (service
  invocation).  Step \texttt{3} corresponds to interaction \texttt{d}
  in Figure~\ref{fig:architectures}(II), step \texttt{2} corresponds
  to interaction \texttt{c$_1$}, step \texttt{4} maps to interaction
  \texttt{b}, and steps 7 and 9 map to \texttt{c$_2$} (in VIP,
  workflow data transfers are performed by the tasks and embedded in
  their descriptions).}
\label{fig:vip-architecture}
\end{figure*}

@@

Task encapsulation is used in the CBRAIN gateway~\cite{SHER-14} to
integrate the FSL toolkit~\cite{Jenkinson2012782} and the PSOM
workflow engine~\cite{bellec2012pipeline}. The CBRAIN-FSL integration,
shown in Figure~\subref*{fig:cbrain-fsl-architecture}, allows
leveraging FSL pipelines written in low-level workflow languages
(Linux executables and scripts) that submit tasks uniformly through a
specific tool called fsl\_sub. The science gateway is actually split
in two services, the CBRAIN portal and the CBRAIN execution server
deployed on a head node of the infrastructure. The mapping of
interactions \texttt{b}, \texttt{c$_1$} and \texttt{e} to the
CBRAIN-FSL interactions is straightforward.

@@

\begin{figure*}
\centering
\subfloat[CBRAIN-FSL integration.]{
\includegraphics[width=\columnwidth]{figures/CBRAIN-FSL.eps}
\label{fig:cbrain-fsl-architecture}
} \hfill \subfloat[CBRAIN-PSOM integration. Figure
adapted from~\cite{GLAT-16}.]{
  \includegraphics[width=\columnwidth]{figures/CBRAIN-PSOM.eps}
\label{fig:cbrain-psom-architecture}
}
\caption{CBRAIN architecture for workflow engine integration
  (task encapsulation).  1: User sends data and workflow execution request to
  storage server(s) and CBRAIN portal. 2: CBRAIN portal sends
  execution request to execution server on cluster. 3: Execution
  server transfers data from storage server(s). 4-5: Execution server
  starts workflow engine (FSL tool or PSOM master) via task
  scheduler. 6: Workflow engine submits sub-tasks to execution server
  (FSL tasks or PSOM agents). 7-8: Execution server starts sub-tasks
  through task scheduler (FSL sub-tasks or PSOM workers). FSL sub-tasks will run locally
  instead of being submitted again to CBRAIN through interaction 6. 8':
  (PSOM only) PSOM master and workers execute workflow. 9. Execution
  server transfers results to storage server(s). Interaction
  \texttt{b} in Figure~\ref{fig:architectures}(III) is implemented by steps
  \texttt{4}, \texttt{5}, \texttt{7} and \texttt{8} (regular
  interactions with batch manager). Interaction \texttt{c} is
  implemented by steps \texttt{3} and \texttt{9}. Interaction
  \texttt{e} is implemented by step \texttt{6} (for FSL: through a
  modified version of the fsl\_sub script available in
  \url{https://github.com/aces/cbrain-plugins-neuro}; for PSOM:
  through a specific development in PSOM).}
\label{fig:cbrain-sub-tasking}
\end{figure*}

@@

Figure~\ref{fig:tavaxy} shows the architecture of Tavaxy where
workflows in various formats (tSCUFL, SCUFL, t2flow and JSON) can be
imported. As detailed in~\cite{Abouelhoda2012}, workflow conversion
(interaction \texttt{g}) is implemented through an interactive
authoring module where users can check and edit workflow imports, and
a mapper module that performs some language substitutions. The
workflow engine is a re-engineered version of the Galaxy engine that
also includes some components of Taverna.
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/Tavaxy.eps}
\caption{\correction{Architecture of Tavaxy (workflow conversion). Figure adapted
  from~\cite{Abouelhoda2012} with permission of the first author}.}
\label{fig:tavaxy}
\end{figure}

@@

The architectures are
evaluated in Table~\ref{table:evaluation} using five main criteria:
integration complexity, robustness, extensibility, scalability and
\correction{functionality}. Criteria break down into
specific metrics where \emph{lower value indicates better
  performance}. Colors in Table~\ref{table:evaluation} represent
normalized scores (see caption). For each criterion, a global score is
computed by summing up the individual scores for each metric. We
ensure that the different metrics used to calculate the global score
for a criterion are comparable, so that they can be combined
meaningfully.

\correction{The metrics defined hereafter aim at providing a framework
  to evaluate architectures independently from any particular system,
  regardless of any technical constraint such as the use of a specific
  programming language or the internal architecture of the science
  gateway. The goal is to abstract architectural discussions from the
  particular technical context of a real system. Some degree of
  system-specificity can be introduced in the evaluation by assigning
  weights to interactions and components through a public spreadsheet
  available at \url{https://frama.link/LrPRsCQp}\footnote{The
    spreadsheet may need to be downloaded to be edited.}. In
  particular, sheet ``Components and interactions'' in this
  spreadsheet contains the weight assigned to components and
  interactions, which can be used to produce new versions of
  Table~\ref{table:evaluation} that fit the particular technical
  context of a real system. For instance, a high weight could be
  assigned to the science gateway if its internal architecture is such
  that it can hardly be extended in practice.}

\correction{In addition, all the metrics in a criterion contribute
  equally to the global score used for the criterion. Column B of
  sheet ``Metrics'' of the public spreadsheet may be used to assign
  different weights to each metric to adjust the specific context of a
  study.

  The remainder of this Section defines the metrics in each criterion.
}

@@

\correction{Integration complexity} is measured through the number of
software components and their interactions. It breaks down into the
following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total components (\texttt{I$_1$}): total number of components in
  the architecture.
\item Total interactions (\texttt{I$_2$}): total number of
  interactions in the architecture.
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$}, since it is usually an external entity that is not
developed nor maintained by the groups who integrate workflow engines
in science gateway. We support its inclusion here because integrating
an infrastructure into the system usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.),
therefore increasing complexity. \correction{The weights given to
  software components in I$_1$ and interactions in I$_2$ can be
  adjusted in the public spreadsheet to cover the technical
  peculiarities of a real system.}

 The two metrics
\correction{I$_1$ and I$_2$} are summed to obtain a global score that
measures the total number of software pieces affected by integration.

@@

\correction{Functionality includes}:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{\correction{F}$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{\correction{F}$_1$}=0 means that meta-workflows are
  intrinsically enabled by the architecture (i.e., they can be
  implemented using the components and interactions already in place),
  and \texttt{\correction{F}$_1$}=1 means that they may be implemented with some
  development effort.

@@

\section{\correction{Architecture evaluations}}

@@

@@

Differences in \correction{functionality} should not be
neglected. Nested workflows (with service invocation) and workflow
conversion (with service invocation) are the only architectures that
intrinsically support meta-workflows. Besides, tight integration is
the only architecture that allows fine-grained debugging, which may be
critical for efficient user support.

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
\correction{It should be noted} that this analysis is only meant for illustration purposes, since
it assumes that all criteria have equal weights while real systems
would favor some of them. \correction{Weights can be adjusted in the
  public spreadsheet at \url{https://frama.link/LrPRsCQp} (column B of
  sheet ``Metrics'') to produce a custom version of
  Table~\ref{table:overall} if required.}
